Icicle: a single-pass query language

### 2015-12-05 2016icicle/section/02-Source.tex
ok

>>>>

This language is not referentially transparent, which may seem odd, but this actually allows us to restrict the language and uphold our performance guarantees.
The input stream is implicitly passed as an argument to all parts of the program, but cannot be referred to explicitly.
Additionally, operations like @filter@ create a new context where the input stream is modified.
This means that a @let@ binding outside of a @filter@ can have different semantics when moved inside the @filter@, despite having the same body.
However, by passing this argument implicitly we remove a class of programs that would be hard to guarantee performance for.
By prohibiting the programmer from accessing the underlying stream, they cannot perform certain operations like zipping two streams with different filters together.

### 2015-12-05 2016icicle/section/03-Types.tex
ok

@@

%%% Types
In figure~\ref{fig:source:type:defs} we define the types.
We use $T$ for base types, of which we only have natural numbers and booleans.

@@

%%% Function application
In figure~\ref{fig:source:type:wrap} there are two judgments defined, which are both used for the type of function application.
In contrast to traditional modality type systems, we infer the boxing and unboxing of various modalities.

@@

The judgment $\WrapMode{\tau}{m}{\tau'}$, corresponds to reboxing the function return type $\tau$ into modality $m$, returning $\tau'$ as the boxed return type.
The rule (WrapModeArrow) simply recurses down through the function arguments, applying the judgment to the result.
The rule (WrapElement) applies when reboxing an @Element@ result as an @Element@, and similarly for (WrapAggregate).
For (WrapPure) the result type is a @Pure@, so it can be boxed to any mode.

The judgment $\WrapApp{\tau}{\phi}{\tau'}$ attempts to apply the function type $\tau$ to argument $\phi$, performing boxing and unboxing if necessary.
If both modalities and types are equal, (WrapApp) applies the function as normal.
When the function expects a @Pure@ argument and is passed an @Element@ or an @Aggregate@, we perform reboxing as in (WrapAppElement) and (WrapAppAggregate).
Finally, no matter what the function expects, if it is passed a @Pure@ it is simply boxed and applied as in (WrapAppPure), with no need to rebox the result.

\TODO{Or should we just use an explicitly box/unbox version?}

%%% Expressions
With function application already defined, typechecking expressions in figure~\ref{fig:source:type:ctx} becomes quite standard.
The judgment $\Typecheck{\Gamma}{x}{\tau}$ means that under a context $\Gamma$, the expression $x$ has type and modality $\tau$.
Being a first-order language, the typing rule returns a first-order type rather than a function type.

Rules (TcVar) and (TcPrim) simply look up the variable or primitive in the environment.

For (TcApp) we simply use the boxed function application judgment from figure~\ref{fig:source:type:wrap}.

\TODO{I think we need a box/cast rule for going from @Pure@ to @Element@ etc}
For simplicity, we reuse the same judgment for expressions and contexts.

@@

\input{figures/Source-Compile.tex}

>>>>

We will now formally describe the typesystem of Icicle, and show how modality types are used to enforce the single-pass restriction.

@@

\subsection{Types}
In figure~\ref{fig:source:type:defs} we define the types.
We use $T$ for base types, of which we only have natural numbers and booleans.

@@

\subsection{Function application with unboxing}
In figure~\ref{fig:source:type:wrap} there are two judgments defined, which are both used for the type of function application.
In contrast to traditional modality type systems, we infer the boxing and unboxing of various modalities.
\TODO{Citation and find something similar.}
\TODO{Probably talk about comonads or something}

@@

These function application judgments only deal with the unboxing case; boxing is dealt with later as the single rule (TcBox).

The judgment $\WrapMode{\tau}{m}{\tau'}$, corresponds to unboxing the a function argument of type $m$ and reboxing the function return type $\tau$ into modality $m$, with $\tau'$ as the boxed return type.
The rule (WrapModeArrow) simply recurses down through the function arguments, applying the judgment to the result.
The rule (WrapElement) applies when reboxing an @Element@ result as an @Element@, and similarly for (WrapAggregate).
For (WrapPure) the result type is a @Pure@, so it can be boxed to any mode.
If these rules do not apply, for example when an @Aggregate@ result is attempted to be reboxed as an @Element@, the function application is an error.

The judgment $\WrapApp{\tau}{\phi\ldots}{\tau'}$ attempts to apply the function type $\tau$ to the argument list $\phi\ldots$, performing unboxing as necessary.
When the argument list is empty and the function type is a nullary function or simple value type, (WrapAppFinished) simply returns the result type.

In the next rules, we use the syntax $a;~b$ for a list with $a$ as its head and $b$ as its tail.

When the actual argument and expected function argument modalities and types are equal, (WrapAppEqual) applies the function as normal, and continues along with any remaining arguments.

When the function expects a @Pure@ argument and is passed an @Element@ or an @Aggregate@, we perform reboxing as in (WrapAppPure).
This attempts to rebox the function result to the argument modality, and then continues by wrapping the remaining arguments.

\subsection{Expression}
With function application already defined, typechecking expressions in figure~\ref{fig:source:type:ctx} becomes quite standard.
The judgment $\Typecheck{\Gamma}{x}{\tau}$ means that under a context $\Gamma$, the expression $x$ has type and modality $\tau$.
Being a first-order language, the typing rule returns a first-order type rather than a function type.

Rules (TcVar) and (TcPrim) simply look up the variable or primitive in the environment.
While the environment contains function types $T_\to$, these rules only apply if the result is a nullary function with no arguments.

The rule (TcBox) allows any @Pure@ computation to be implicitly cast to another mode, such as @Element@ or @Aggregate@.
This is equivalent to boxing the expression.

There are two different application rules (TcAppVar) and (TcAppPrim), which simply look up the function type in different environments before using the boxed function application judgement from figure~\ref{fig:source:type:wrap}.
In these ruler we use ``$\ldots$'' as shorthand for finding the types of all arguments.

@@

\subsection{Single-pass restriction}

We now describe how this typesystem encodes the single-pass restriction: that is, that if a given query type-checks, it only needs to look at each element in the stream once.

First, look at an example that would require two passes over the data: maybe @filter open > mean open in count@.
In this case, $\mit{open}$ has type $\Elm{\NN}$ while $\mit{mean~open}$ has type $\Agg{\NN}$.
The primitive $>$ has type $\NN \to \NN \to \NN$ and can be boxed to either @Element@ or @Pure@, but not both.
This means that $\mit{open} > \mit{mean open}$ is a type error.

In Icicle, folding and boxing pure values are the only ways to produce @Aggregate@ types.
The other typing rules that refer to @Aggregate@s require an existing @Aggregate@ and do not produce an aggregate on their own.
This means that all @Aggregate@s must either be the result of folds, which in general require all their input before they can be computed, or pure computations.

There is also no way to take an @Aggregate@ and convert it to a @Pure@ or an @Element@, except for the @scan@ primitive which explicitly uses a fold's intermediate value, rather than the value of the entire pass.

The only way to require multiple passes over the data, therefore, is to have a @Pure@ or @Element@ computation that depends on the result of an @Aggregate@ computation.
And so on.
